Collective excitations in competing phases in two and three dimensions

### 2024-03-15 paper/main.tex
Feierabend

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                  Abstract                                                 %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
    We investigate the superconducting (SC), charge-density wave (CDW), and antiferromagnetic (AFM) phases in the extended Hubbard model at zero temperature and half-filling. 
    We employ the iterated equations of motion approach to compute the two-particle Green's functions and their spectral densities. 
    This renders a comprehensive analysis of the behavior of collective excitations possible as the model's parameters are changed across phase transitions. 
    We identify the well-known amplitude (Higgs) and phase (Anderson-Bogoliubov) modes within the superconducting phase 
    and observe a similar excitation (cooperon) in the CDW phase which shifts towards zero energy as the system approaches the phase transition to the SC phase. 
    \textcolor{red}{In the CDW phase, close to the phase transition to the AFM phase, 
    we find a collective mode (exciton) that does not change significantly and another mode (magnon) that moves into the two-particle continuum. 
    In the AFM phase, we identify both longitudinal and transversal magnons.}
\end{abstract}

@@

Numerous studies have investigated the phases and various quantities of the Hubbard model in equilibrium systems, including various additional interactions with and without doping.
\cite{Micnas88,Micnas88b,Micnas89,Dzierzawa92,Kostyrko92,Eriksson95,Staudt00,Onari04,Toschi05,Brackett16,Paki19,romer20,Sushchyev22}.
\textcolor{red}{Recent studies on the dynamics of the superconducting gap parameter examined quenches, yielding oscillations \cite{Volkov73,Yuzbashyan05,Yuzbashyan06,Barankov06,Cui19}, 
and driven systems in the hope of inducing superconductivity \cite{Nicoletti14,Krull14,Moor14,Casandruc15,patel16,sentef17,Buenemann17}.}

In this paper, we will restrict ourselves to the half-filled Hubbard model, including an additional intersite interaction, on a square and a simple cubic lattice.
\textcolor{red}{Specifically, we investigate competing phases, i.e., the behavior of various collective modes around phase transitions.
To this end, the 2D case already provides a wide range of possible phases, including CDW, AFM, $s$- and $d_{x^2 - y^2}$-wave superconductivity, and phase-separated states \cite{Micnas88b,Tsuchiura95,Su01,Su04,ha11,Huang13,Jiang22,Linner23}.}
We expand these well-known phase diagrams to the simple cubic lattice where we find a similar structure for CDW, AFM, and $s$-wave SC.

@@

Moreover, we demonstrate an explicit way to obtain various Green's functions from the aforementioned methodology.
By extension, we also obtain the spectral functions of the investigated systems and discuss the excitations found therein.
The prominent examples in the SC phase are the well-known phase (Anderson-Bogoliubov) and amplitude (Higgs) modes.
\textcolor{red}{The former occurs in neutral superfluids at $0$ energy and has been found in the past \cite{Bogoljubov1958,Anderson58,Brieskorn74,Schmid1975,simanek1975,schoen1976propagating,Maiti2015,Sun2020,Fan22}.
We will obtain this mode as well based on a microscopic description that excludes long-range electromagnetic interactions.
The inclusion of this kind of interaction would shift this mode towards the plasma frequency \cite{Anderson58,schoen1976propagating,Kulik1981}.}
\textcolor{red}{The amplitude mode, on the other hand, is located at the lower edge of the quasiparticle continuum as it represents the energy required to form or break a Cooper pair \cite{Schmid1975,Varma02,Cea14,Measson14,Tsuji15,Krull16,Mueller2019,Schwarz20}}

@@

\textcolor{red}{We will investigate this model for various sets of parameters allowing us access to an assortment of phases.
Namely, the model exhibits antiferromagnetism for positive $U$ and moderately sized $V$.
At larger $V$ the intersite repulsion takes over and favors a charge-ordered phase.
In the $U<0$ region, the same phase occurs for any $V>0$.
Choosing $V=0$ leads to a coexistence of $s$-wave superconductivity and said CDW \cite{yang90}, while moderately sized $V<0$ yield a purely $s$-wave superconducting phase \cite{Micnas88b}.
In this article, we will explore the AFM-CDW as well as the CDW-SC phase transitions.}

@@

\begin{figure}
    \centering
    \includegraphics[width=.48\textwidth]{plots/phase_diagram.pdf}
    \caption{The phase diagram obtained for the extended Hubbard model on 
    (a) a square lattice and (b) a simple cubic lattice using a mean-field approximation at temperature $T=0$.
    The CDW-AFM boundary lies at $U = zV$, where $z$ is the coordination number. 
    For $U<0$ and $V=0$, there is a coexistence of CDW and $s$-wave SC.
    The boundary for the $d_{x^2 - y^2}$-wave SC (dashed line) for the square lattice has been taken from Ref. \cite{Micnas88b}, 
    as it is not accessible using our formalism.
    \textcolor{red}{The red shading shows the region in which the dynamical matrix $\mM$, see equation \eqref{eqn:heisenberg}, has negative eigenvalues, 
    indicating that there are phases not captured by our methods.
    In accordance with the literature, we expect to find a phase-separated state \cite{Linner23}.
    Lastly, the striped area shows the region that we cannot access because the discretization $N_\gamma$, 
    which is required to accurately describe the system, is too large. However, we expect to find a superconducting state here.}}
    \label{fig:phase_diagram}
\end{figure}

@@

In this article, we will be using the \textcolor{red}{symplectic product,}
\begin{equation}
\label{eqn:scalar_product}
    (A | B) \coloneqq  \langle [A^\dagger, B] \rangle\,,
\end{equation}
where the expectation values are taken with respect to the thermal equilibrium of the mean-field Hamiltonian \eqref{eqn:mf_hamiltonian}.
Note, that this is not a proper scalar product, as a norm defined using it can also be negative.
For example, assume that for some operator $A$
\begin{equation}
    ||A|| = (A | A) = \langle [A^\dagger, A] \rangle > 0\,,
\end{equation}
then
\begin{equation}
    ||A^\dagger|| = (A^\dagger | A^\dagger) = \langle [A, A^\dagger] \rangle = - ||A|| < 0\,.
\end{equation}
Nevertheless, this does not pose any issue in our calculations.
We will compute $\mM$ by commutating with the full Hubbard Hamiltonian \eqref{eqn:full_hamiltonian} enabling us to capture collective behavior rather than the one particle dynamics given by the mean-field approximation.

\textcolor{red}{To treat the square and simple cubic lattice on equal footing, we utilize the fact \eqref{eqn:equal_expecs} to create our operator set in $\gamma$-space.}
It consists of operators of the type
\begin{equation}
    \label{eqn:ieom_basis_operator}
    A_\gamma \coloneqq \frac{1}{\sqrt{N}} \sum_{\vk} \delta (\gamma - \hat{\gamma}( \vk )) A_{\vk}\,,
\end{equation}
where $A_{\vk}$ represents each type of operator in \eqref{eqn:operators} and their Hermitian conjugates.
Operators with different indices are put into the set individually, e.g., $n_{\gamma \up}$, and $n_{\gamma \down}$ are distinct operators in our set.
\textcolor{red}{Additionally, we will include an analogous expression for
\begin{equation}
    \tau_{k} \coloneqq c_{k \uparrow}^\dagger c_{k \downarrow}
\end{equation}
and its Hermitian conjugate.
This will allow us to describe transversal magnons later on.}

Naturally, we have to discretize $\gamma$, 
however, this allows us to obtain good results with drastically smaller matrices compared to a momentum-based operator set and obtain results for a three-dimensional system in the first place.
Practically, we choose $N_\gamma = 6000$ equally spaced discretization points.
A detailed explanation of the numerics is provided in the appendix \ref{sec:numerical_ieom}.
\textcolor{red}{Computationally, our method is limited by the number of terms included in our operator set as well in the magnitude of $N_\gamma$.}
If the gap $\Delta_\text{tot}$ is of the same order of magnitude as our discretization $\Delta \gamma$ our numerics become inaccurate.
This manifests specifically if a spectral function has strong features at minuscule energies, e.g., the phase mode in the superconducting phase.
In this case, the dynamical matrix $\mM$ picks up negative eigenvalues that vanish if $N_\gamma$ is increased so that $\Delta \gamma$ is sufficiently smaller than $\Delta_\text{tot}$.
This happens if the parameters are chosen from the striped area in \autoref{fig:phase_diagram}.
\textcolor{red}{This issue only arises in the SC phase, nevertheless, we expect similar inaccuracies in the AFM region for small $U$.
However, we did not perform any computations in that region, hence this does not pose an issue to our analysis.}

@@

Due to this choice of set, the matrices occurring in \eqref{eqn:heisenberg} have a special block structure
\begin{equation}
    \label{eqn:xp_set}
    \mM = \begin{pmatrix}
        \mathcal{K}_+ & \kappa \\ \kappa^\dagger & \mathcal{K}_-
    \end{pmatrix}\,,\quad \mN = \begin{pmatrix}
        \Lambda_+ & \mathcal{L} \\ \mathcal{L}^\dagger & \Lambda_-
    \end{pmatrix}\,,
\end{equation}
where the upper block refers to all $X_i$ and the lower block to all $P_i$.
Furthermore, due to symmetry, the relations $\Im [\mathcal{K}_\pm] = \Re [\kappa] = 0$ and $\Re [\Lambda_\pm] = \Im [\mathcal{L}] = 0$ hold.
Thus, if all occurring expectation values are real, both matrices have large empty blocks.
This condition is fulfilled for all cases investigated in this article \textcolor{red}{because our model has inversion symmetry as well as time-reversal symmetry.}

To study the collective excitations, we investigate various retarded Green's functions 
\begin{equation}
    G_{AB}^\text{ret} (t) = - \im \langle [A(t), B] \rangle \Theta(t)\,,
\end{equation}
and their Fourier-transformed
\begin{equation}
    \label{eqn:standard_gf}
    G_{AB}(z = \omega + \im 0^+) = -\im \int_0^{\infty} e^{\im z t} \langle [A(t), B] \rangle \mathrm{d}t\,,
\end{equation}
where $\Theta(t)$ is the Heaviside function.
Our previous choice of pseudo scalar product enables us to write down a matrix-valued Green's functions in terms of the aforementioned matrices
\begin{align}
    \label{eqn:green_function}
    \mathcal{G}(z &= \omega + \im 0^+) = \mN \frac{1}{-z \mN - \mM} \mN \\
        &\coloneqq  -\mN \mathcal{R}(z) \mN\,,
\end{align}
where we used the resolvent
\begin{equation}
    \label{eqn:resolvent}
    \mathcal{R}(z) \coloneqq  \frac{1}{z \mN + \mM}\,.
\end{equation}
The entries of this matrix $\mathcal{G}(z)$ are the Fourier-transformed Green's functions mentioned in \eqref{eqn:standard_gf} with respect to the operators from the initially chosen set.
\textcolor{red}{That means, if we index each operator in our set $A_i$, the matrix element $(j,i)$ of $\mathcal{G}$ is the Green's function
\begin{align}
    \mathcal{G}_{ji}(z = \omega +\im 0^+) &=  G_{A_j A_i^\dagger} (z) \nonumber \\
        &= -\im \int_0^{\infty} \langle [A_j(t), A_i^\dagger(0)] \rangle e^{\im z t} \mathrm{d}t\,.
\end{align}}

In the following, we will derive this relation.
\textcolor{red}{For that, consider the Fourier-transformed Green's function above.
We rewrite both operators in the same manner as in \eqref{eqn:time_dependent_operator} to obtain
\begin{align}
    G_{AB^\dagger} (z) &= \im \int_0^\infty e^{\im z t} \langle [ A_i^\dagger(0), A_j(t) ] \rangle \mathrm{d}t \nonumber \\
        &= \im \int_0^\infty e^{\im z t} \langle [ \sum_m a_{j,m}^*(0) A_m^\dagger, \sum_n a_{i,n}(t) A_n ] \rangle \mathrm{d}t \nonumber \\
        &= \im \int_0^\infty \sum_{mn} e^{\im z t} a_{j,m}^*(0) \underbrace{\langle [ A_m^\dagger, A_n ] \rangle}_{\equiv \mN_{mn}} a_{i,n}(t) \mathrm{d}t \nonumber \\
        &= \im \int_0^\infty e^{\im z t} \vec{a}_{j}^\dagger (0) \mN \vec{a}_{i}(t) \mathrm{d}t\,.
\end{align}
Furthermore, assuming, that $\mN$ is invertible, we can easily solve the differential equation \eqref{eqn:heisenberg} to obtain
\begin{equation}
    \vec{a}_{i}(t) = \exp \left( \im \mN^{-1} \mM t \right) \vec{a}_{i}(0)\,.
\end{equation}
This allows us to finally obtain
\begin{align}
    \label{eqn:green_derivation}
    G_{AB^\dagger} (z) &= \im  \int_0^\infty \vec{a}_{j}^\dagger (0) \mN \exp \left( \im \left(\mN^{-1} \mM + z \right) t \right) \vec{a}_{i}(0) \mathrm{d}t  \nonumber \\
        &= - \vec{a}_{j}^\dagger (0) \left[ \mN \frac{1}{\mN^{-1} \mM + z} \right] \vec{a}_{i}(0) \nonumber \\
        &= - \vec{a}_{j}^\dagger (0) \left[ \mN \frac{1}{ \underbrace{\mM + z \mN}}_{\equiv \mathcal{R}(z)} \mN \right] \vec{a}_{i}(0)\,,
\end{align}
where $\vec{a}_i(0)$ and $\vec{a}_j^\dagger (0)$ need to be chosen such, that $A_i = \sum_n a_{i,n} A_n$ and $A_j^\dagger = \sum_n a_{j,n}^* A_n^\dagger$ hold.
The task at hand is finding the resolvent as $\mathcal{R}$.}

To do that, we embrace this matrix structure in \eqref{eqn:xp_set} and assume that all matrix entries are real \textcolor{red}{so that $\kappa_ij = \Lambda_{\pm,ij} = 0$ holds}.
Then, with $\mathcal{R}|_X$ denoting the block of a matrix $\mathcal{R}$ corresponding to the $X$-operators, we can calculate
\begin{align}
    r_X (z) &\coloneqq \mathcal{R}(z)|_X = \left. \frac{1}{\mM + z \mN} \right\vert_X \nonumber \\
        &= \left[ \frac{1}{\mM} - \frac{1}{\mM} z \mN \frac{1}{\mM} + \frac{1}{\mM} z \mN \frac{1}{\mM} z \mN \frac{1}{\mM} - \cdots \right]_X \nonumber \\
        &= \left. \frac{1}{\mM} \sum_{j=0}^\infty \left( -z \mN \frac{1}{\mM} \right)^j \right\vert_X \,.
\end{align}
We may omit every second element of the sum, as we want to obtain the upper left block of the matrix, and $\mN$ always swaps the upper left and lower right block.
Thus, we obtain
\begin{align}
    r_X (z) &= \left. \frac{1}{\mM} \sum_{j=0}^\infty z^{2j} \left( \mN \frac{1}{\mM} \right)^{2j} \right\vert_X \nonumber \\
        &= \left. \frac{1}{\mM - z^2 \mN \mM^{-1} \mN} \right\vert_X \nonumber \\
        &= \frac{1}{\mathcal{K}_+ - z^2 \mathcal{L} \mathcal{K}_-^{-1} \mathcal{L}^\dagger}\,.
\end{align}
Next, we define $\check{N} \coloneqq \mathcal{L} \mathcal{K}_-^{-1} \mathcal{L}^\dagger$ for brevity.
Since $\mathcal{M}$ is positive semidefinite and block diagonal, its blocks $\mathcal{K}_\pm$ must be positive semidefinite as well.
Therefore, $\check{N}$ is also positive semidefinite and we can define its square root.
We obtain
\begin{equation}
    \label{eqn:rx}
    r_X (z) = -\check{N}^{-1/2} \frac{1}{z^2 - \check{N}^{-1/2} \mathcal{K}_+ \check{N}^{-1/2}} \check{N}^{-1/2}\,.
\end{equation}
The matrix $\check{M} \coloneqq \check{N}^{-1/2} \mathcal{K}_+ \check{N}^{-1/2}$ is positive semidefinite and Hermitian by the previous arguments.
If $\check{N}$ has a singular part, the inverse can be replaced by the Moore-Penrose pseudo-inverse.
\textcolor{red}{A possible way to compute this is to diagonalize $\check{N} = \mathcal{V} \mathcal{D} \mathcal{V}^\dagger$, with a diagonal matrix $\mathcal{D}$ and a unitary transformation matrix $\mathcal{V}$.
Now, the pseudo-inverse can be defined as
\begin{equation}
    \check{N}^+ \coloneqq \mathcal{V} \mathcal{D}^+ \mathcal{V}^\dagger
\end{equation}
with
\begin{equation}
    (\mathcal{D}^+)_{ij} = \begin{cases}
        1/\mathcal{D}_{ij} & \mathcal{D}_{ij} \neq 0 \\ 0 & \text{otherwise}
    \end{cases}\,.
\end{equation}
Note, that this definition yields the standard inverse for invertible matrices.}

>>>>

\newcommand{\markEdited}{red}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                  Abstract                                                 %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
    We investigate the superconducting (SC), charge-density wave (CDW), and antiferromagnetic (AFM) phases in the extended Hubbard model at zero temperature and half-filling. 
    We employ the iterated equations of motion approach to compute the two-particle Green's functions and their spectral densities. 
    This renders a comprehensive analysis of the behavior of collective excitations possible as the model's parameters are changed across phase transitions. 
    We identify the well-known amplitude (Higgs) and phase (Anderson-Bogoliubov) modes within the superconducting phase 
    and observe a similar excitation (cooperon) in the CDW phase which shifts towards zero energy as the system approaches the phase transition to the SC phase. 
    \textcolor{\markEdited}{In the CDW phase, close to the phase transition to the AFM phase, 
    we find a collective mode (exciton) that does not change significantly and another mode (magnon) that moves into the two-particle continuum. 
    In the AFM phase, we identify both longitudinal and transversal magnons.}
\end{abstract}

@@

Numerous studies have investigated the phases and various quantities of the Hubbard model in equilibrium systems, including various additional interactions with and without doping.
\cite{Micnas88,Micnas88b,Micnas89,Dzierzawa92,Kostyrko92,Eriksson95,Staudt00,Onari04,Toschi05,Brackett16,Paki19,romer20,Sushchyev22}.
\textcolor{\markEdited}{Recent studies on the dynamics of the superconducting gap parameter examined quenches, yielding oscillations \cite{Volkov73,Yuzbashyan05,Yuzbashyan06,Barankov06,Cui19}, 
and driven systems in the hope of inducing superconductivity \cite{Nicoletti14,Krull14,Moor14,Casandruc15,patel16,sentef17,Buenemann17}.}

In this paper, we will restrict ourselves to the half-filled Hubbard model, including an additional intersite interaction, on a square and a simple cubic lattice.
\textcolor{\markEdited}{Specifically, we investigate competing phases, i.e., the behavior of various collective modes around phase transitions.
To this end, the 2D case already provides a wide range of possible phases, including CDW, AFM, $s$- and $d_{x^2 - y^2}$-wave superconductivity, and phase-separated states \cite{Micnas88b,Tsuchiura95,Su01,Su04,ha11,Huang13,Jiang22,Linner23}.}
We expand these well-known phase diagrams to the simple cubic lattice where we find a similar structure for CDW, AFM, and $s$-wave SC.

@@

Moreover, we demonstrate an explicit way to obtain various Green's functions from the aforementioned methodology.
By extension, we also obtain the spectral functions of the investigated systems and discuss the excitations found therein.
The prominent examples in the SC phase are the well-known phase (Anderson-Bogoliubov) and amplitude (Higgs) modes.
\textcolor{\markEdited}{The former occurs in neutral superfluids at $0$ energy and has been found in the past \cite{Bogoljubov1958,Anderson58,Brieskorn74,Schmid1975,simanek1975,schoen1976propagating,Maiti2015,Sun2020,Fan22}.
We will obtain this mode as well based on a microscopic description that excludes long-range electromagnetic interactions.
The inclusion of this kind of interaction would shift this mode towards the plasma frequency \cite{Anderson58,schoen1976propagating,Kulik1981}.}
\textcolor{\markEdited}{The amplitude mode, on the other hand, is located at the lower edge of the quasiparticle continuum as it represents the energy required to form or break a Cooper pair \cite{Schmid1975,Varma02,Cea14,Measson14,Tsuji15,Krull16,Mueller2019,Schwarz20}}

@@

\textcolor{\markEdited}{We will investigate this model for various sets of parameters allowing us access to an assortment of phases.
Namely, the model exhibits antiferromagnetism for positive $U$ and moderately sized $V$.
At larger $V$ the intersite repulsion takes over and favors a charge-ordered phase.
In the $U<0$ region, the same phase occurs for any $V>0$.
Choosing $V=0$ leads to a coexistence of $s$-wave superconductivity and said CDW \cite{yang90}, while moderately sized $V<0$ yield a purely $s$-wave superconducting phase \cite{Micnas88b}.
In this article, we will explore the AFM-CDW as well as the CDW-SC phase transitions.}

@@

\begin{figure}
    \centering
    \includegraphics[width=.48\textwidth]{plots/phase_diagram.pdf}
    \caption{The phase diagram obtained for the extended Hubbard model on 
    (a) a square lattice and (b) a simple cubic lattice using a mean-field approximation at temperature $T=0$.
    The CDW-AFM boundary lies at $U = zV$, where $z$ is the coordination number. 
    For $U<0$ and $V=0$, there is a coexistence of CDW and $s$-wave SC.
    The boundary for the $d_{x^2 - y^2}$-wave SC (dashed line) for the square lattice has been taken from Ref. \cite{Micnas88b}, 
    as it is not accessible using our formalism.
    \textcolor{\markEdited}{The red shading shows the region in which the dynamical matrix $\mM$, see equation \eqref{eqn:heisenberg}, has negative eigenvalues, 
    indicating that there are phases not captured by our methods.
    In accordance with the literature, we expect to find a phase-separated state \cite{Linner23}.
    Lastly, the striped area shows the region that we cannot access because the discretization $N_\gamma$, 
    which is required to accurately describe the system, is too large. However, we expect to find a superconducting state here.}}
    \label{fig:phase_diagram}
\end{figure}

@@

In this article, we will be using the \textcolor{\markEdited}{symplectic product,}
\begin{equation}
\label{eqn:scalar_product}
    (A | B) \coloneqq  \langle [A^\dagger, B] \rangle\,,
\end{equation}
where the expectation values are taken with respect to the thermal equilibrium of the mean-field Hamiltonian \eqref{eqn:mf_hamiltonian}.
Note, that this is not a proper scalar product, as a norm defined using it can also be negative.
For example, assume that for some operator $A$
\begin{equation}
    ||A|| = (A | A) = \langle [A^\dagger, A] \rangle > 0\,,
\end{equation}
then
\begin{equation}
    ||A^\dagger|| = (A^\dagger | A^\dagger) = \langle [A, A^\dagger] \rangle = - ||A|| < 0\,.
\end{equation}
Nevertheless, this does not pose any issue in our calculations.
We will compute $\mM$ by commutating with the full Hubbard Hamiltonian \eqref{eqn:full_hamiltonian} enabling us to capture collective behavior rather than the one particle dynamics given by the mean-field approximation.

\textcolor{\markEdited}{To treat the square and simple cubic lattice on equal footing, we utilize the fact \eqref{eqn:equal_expecs} to create our operator set in $\gamma$-space.}
It consists of operators of the type
\begin{equation}
    \label{eqn:ieom_basis_operator}
    A_\gamma \coloneqq \frac{1}{\sqrt{N}} \sum_{\vk} \delta (\gamma - \hat{\gamma}( \vk )) A_{\vk}\,,
\end{equation}
where $A_{\vk}$ represents each type of operator in \eqref{eqn:operators} and their Hermitian conjugates.
Operators with different indices are put into the set individually, e.g., $n_{\gamma \up}$, and $n_{\gamma \down}$ are distinct operators in our set.
\textcolor{\markEdited}{Additionally, we will include an analogous expression for
\begin{equation}
    \tau_{k} \coloneqq c_{k \uparrow}^\dagger c_{k \downarrow}
\end{equation}
and its Hermitian conjugate.
This will allow us to describe transversal magnons later on.}

Naturally, we have to discretize $\gamma$, 
however, this allows us to obtain good results with drastically smaller matrices compared to a momentum-based operator set and obtain results for a three-dimensional system in the first place.
Practically, we choose $N_\gamma = 6000$ equally spaced discretization points.
A detailed explanation of the numerics is provided in the appendix \ref{sec:numerical_ieom}.
\textcolor{\markEdited}{Computationally, our method is limited by the number of terms included in our operator set as well in the magnitude of $N_\gamma$.}
If the gap $\Delta_\text{tot}$ is of the same order of magnitude as our discretization $\Delta \gamma$ our numerics become inaccurate.
This manifests specifically if a spectral function has strong features at minuscule energies, e.g., the phase mode in the superconducting phase.
In this case, the dynamical matrix $\mM$ picks up negative eigenvalues that vanish if $N_\gamma$ is increased so that $\Delta \gamma$ is sufficiently smaller than $\Delta_\text{tot}$.
This happens if the parameters are chosen from the striped area in \autoref{fig:phase_diagram}.
\textcolor{\markEdited}{This issue only arises in the SC phase, nevertheless, we expect similar inaccuracies in the AFM region for small $U$.
However, we did not perform any computations in that region, hence this does not pose an issue to our analysis.}

@@

Due to this choice of set, the matrices occurring in \eqref{eqn:heisenberg} have a special block structure
\begin{equation}
    \label{eqn:xp_set}
    \mM = \begin{pmatrix}
        \mathcal{K}_+ & \kappa \\ \kappa^\dagger & \mathcal{K}_-
    \end{pmatrix}\,,\quad \mN = \begin{pmatrix}
        \Lambda_+ & \mathcal{L} \\ \mathcal{L}^\dagger & \Lambda_-
    \end{pmatrix}\,,
\end{equation}
where the upper block refers to all $X_i$ and the lower block to all $P_i$.
Furthermore, due to symmetry, the relations $\Im [\mathcal{K}_\pm] = \Re [\kappa] = 0$ and $\Re [\Lambda_\pm] = \Im [\mathcal{L}] = 0$ hold.
Thus, if all occurring expectation values are real, both matrices have large empty blocks.
This condition is fulfilled for all cases investigated in this article \textcolor{\markEdited}{because our model has inversion symmetry as well as time-reversal symmetry.}

To study the collective excitations, we investigate various retarded Green's functions 
\begin{equation}
    G_{AB}^\text{ret} (t) = - \im \langle [A(t), B] \rangle \Theta(t)\,,
\end{equation}
and their Fourier-transformed
\begin{equation}
    \label{eqn:standard_gf}
    G_{AB}(z = \omega + \im 0^+) = -\im \int_0^{\infty} e^{\im z t} \langle [A(t), B] \rangle \mathrm{d}t\,,
\end{equation}
where $\Theta(t)$ is the Heaviside function.
Our previous choice of pseudo scalar product enables us to write down a matrix-valued Green's functions in terms of the aforementioned matrices
\begin{align}
    \label{eqn:green_function}
    \mathcal{G}(z &= \omega + \im 0^+) = \mN \frac{1}{-z \mN - \mM} \mN \\
        &\coloneqq  -\mN \mathcal{R}(z) \mN\,,
\end{align}
where we used the resolvent
\begin{equation}
    \label{eqn:resolvent}
    \mathcal{R}(z) \coloneqq  \frac{1}{z \mN + \mM}\,.
\end{equation}
The entries of this matrix $\mathcal{G}(z)$ are the Fourier-transformed Green's functions mentioned in \eqref{eqn:standard_gf} with respect to the operators from the initially chosen set.
\textcolor{\markEdited}{That means, if we index each operator in our set $A_i$, the matrix element $(j,i)$ of $\mathcal{G}$ is the Green's function
\begin{align}
    \mathcal{G}_{ji}(z = \omega +\im 0^+) &=  G_{A_j A_i^\dagger} (z) \nonumber \\
        &= -\im \int_0^{\infty} \langle [A_j(t), A_i^\dagger(0)] \rangle e^{\im z t} \mathrm{d}t\,.
\end{align}}

In the following, we will derive this relation.
\textcolor{\markEdited}{For that, consider the Fourier-transformed Green's function above.
We rewrite both operators in the same manner as in \eqref{eqn:time_dependent_operator} to obtain
\begin{align}
    G_{A_j A_i^\dagger} (z) &= \im \int_0^\infty e^{\im z t} \langle [ A_i^\dagger(0), A_j(t) ] \rangle \mathrm{d}t \nonumber \\
        &= \im \int_0^\infty e^{\im z t} \langle [ \sum_m a_{i,m}^*(0) A_m^\dagger, \sum_n a_{j,n}(t) A_n ] \rangle \mathrm{d}t \nonumber \\
        &= \im \int_0^\infty \sum_{mn} e^{\im z t} a_{i,m}^*(0) \underbrace{\langle [ A_m^\dagger, A_n ] \rangle}_{\equiv \mN_{mn}} a_{j,n}(t) \mathrm{d}t \nonumber \\
        &= \im \int_0^\infty e^{\im z t} \vec{a}_{i}^\dagger (0) \mN \vec{a}_{j}(t) \mathrm{d}t\,.
\end{align}
Furthermore, assuming, that $\mN$ is invertible, we can easily solve the differential equation \eqref{eqn:heisenberg} to obtain
\begin{equation}
    \vec{a}_{j}(t) = \exp \left( \im \mN^{-1} \mM t \right) \vec{a}_{j}(0)\,.
\end{equation}
This allows us to finally obtain
\begin{align}
    \label{eqn:green_derivation}
    G_{A_j A_i^\dagger} (z) &= \im  \int_0^\infty \vec{a}_{i}^\dagger (0) \mN \exp \left( \im \left(\mN^{-1} \mM + z \right) t \right) \vec{a}_{j}(0) \mathrm{d}t  \nonumber \\
        &= - \vec{a}_{i}^\dagger (0) \left[ \mN \frac{1}{\mN^{-1} \mM + z} \right] \vec{a}_{j}(0) \nonumber \\
        &= - \vec{a}_{i}^\dagger (0) \left[ \mN \frac{1}{ \underbrace{\mM + z \mN}}_{\equiv \mathcal{R}(z)} \mN \right] \vec{a}_{j}(0)\,,
\end{align}
where $\vec{a}_j(0)$ and $\vec{a}_i^\dagger (0)$ need to be chosen such, that $A_j = \sum_n a_{j,n} A_n$ and $A_i^\dagger = \sum_n a_{i,n}^* A_n^\dagger$ hold.
The task at hand is finding the resolvent as $\mathcal{R}$.}

To do that, we embrace this matrix structure in \eqref{eqn:xp_set} and assume that all matrix entries are real \textcolor{\markEdited}{so that $\kappa_{ij} = \Lambda_{\pm,ij} = 0$ holds}.
Then, with $\mathcal{R}|_X$ denoting the block of a matrix $\mathcal{R}$ corresponding to the $X$-operators, we can calculate
\begin{align}
    r_X (z) &\coloneqq \mathcal{R}(z)|_X = \left. \frac{1}{\mM + z \mN} \right\vert_X \nonumber \\
        &= \left[ \frac{1}{\mM} - \frac{1}{\mM} z \mN \frac{1}{\mM} + \frac{1}{\mM} z \mN \frac{1}{\mM} z \mN \frac{1}{\mM} - \cdots \right]_X \nonumber \\
        &= \left. \frac{1}{\mM} \sum_{j=0}^\infty \left( -z \mN \frac{1}{\mM} \right)^j \right\vert_X \,.
\end{align}
We may omit every second element of the sum, as we want to obtain the upper left block of the matrix, and $\mN$ always swaps the upper left and lower right block.
Thus, we obtain
\begin{align}
    r_X (z) &= \left. \frac{1}{\mM} \sum_{j=0}^\infty z^{2j} \left( \mN \frac{1}{\mM} \right)^{2j} \right\vert_X \nonumber \\
        &= \left. \frac{1}{\mM - z^2 \mN \mM^{-1} \mN} \right\vert_X \nonumber \\
        &= \frac{1}{\mathcal{K}_+ - z^2 \mathcal{L} \mathcal{K}_-^{-1} \mathcal{L}^\dagger}\,.
\end{align}
Next, we define $\check{N} \coloneqq \mathcal{L} \mathcal{K}_-^{-1} \mathcal{L}^\dagger$ for brevity.
Since $\mathcal{M}$ is positive semidefinite and block diagonal, its blocks $\mathcal{K}_\pm$ must be positive semidefinite as well.
Therefore, $\check{N}$ is also positive semidefinite and we can define its square root.
We obtain
\begin{equation}
    \label{eqn:rx}
    r_X (z) = -\check{N}^{-1/2} \frac{1}{z^2 - \check{N}^{-1/2} \mathcal{K}_+ \check{N}^{-1/2}} \check{N}^{-1/2}\,.
\end{equation}
The matrix $\check{M} \coloneqq \check{N}^{-1/2} \mathcal{K}_+ \check{N}^{-1/2}$ is positive semidefinite and Hermitian by the previous arguments.
If $\check{N}$ has a singular part, the inverse can be replaced by the Moore-Penrose pseudo-inverse.
\textcolor{\markEdited}{A possible way to compute this is to diagonalize $\check{N} = \mathcal{V} \mathcal{D} \mathcal{V}^\dagger$, with a diagonal matrix $\mathcal{D}$ and a unitary transformation matrix $\mathcal{V}$.
Now, the pseudo-inverse can be defined as
\begin{equation}
    \check{N}^+ \coloneqq \mathcal{V} \mathcal{D}^+ \mathcal{V}^\dagger
\end{equation}
with
\begin{equation}
    (\mathcal{D}^+)_{ij} = \begin{cases}
        1/\mathcal{D}_{ij} & \mathcal{D}_{ij} \neq 0 \\ 0 & \text{otherwise}
    \end{cases}\,.
\end{equation}
Note, that this definition yields the standard inverse for invertible matrices.}
